---
title: "Using Biogeographical Stochastic Mapping on Herodotools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using Biogeographical Stochastic Mapping on Herodotools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```



To evaluate the imprints of historical processes on assemblages `{Herodotools}` has based its assessments of historical events on the ancestral area reconstruction. Here, we use the Biogeographical Stochastic Mapping to account for uncertainty in the ancestral area reconstruction as well as to provide a more precise measurements of the timing of the events.


The Biogeographical Stochastic Mapping (BSM) is a method implemented in the `{BioGeoBEARS}` package [(Dupin et al. 2017)](https://doi.org/10.1111/jbi.12898), which uses the parameters of a biogeographical model (e.g. DEC model) to map on the phylogenetic tree the ancestral area on cladogenetic and anagenetic events. The function `runBSM` in the `{BioGeoBEARS}` package uses a probabilistic model of dispersal and vicariance to map the evolution of ancestral range along the branches of the tree. Each map simulates the range evolution on the phylogenetic tree by randomly assigning range area according to the model’s parameters and the phylogeny. It allows you to observe the variability and uncertainty of biogeographic history across multiple runs.

We have included BSM into the `{Herodotools}` workflow and this article shows you how to use the new functionality. 

## Akodon data

We will use data from the genus *Akodon* to exemplify the workflow. In the article **Using Herodotools to analyses of historical biogeography**, we have defined the evolutionary regions, and estimated the biogeographical ancestral area. Here we will use those data to go further and run the BSM. After having the BSM results, we will walk you thought the steps to get calculate the assemblage metrics of in situ diversification, age of arrival, Phylogenetic Endemism and Phylogenetic Diversity, as well as the historical dispersal events from the assemblage level perspective.


## Run Biogeographical Stochastic Mapping

```{r load, message=FALSE, warning=FALSE}

# load the package -------------------------------------------------------------
library(Herodotools)
library(rnaturalearth)
library(sf)
library(rcartocolor)
library(ggplot2)
library(patchwork)


# load the data ----------------------------------------------------------------

# phylogeny
data("akodon_newick")

# Assemblages
data("akodon_sites")

# Biogeographical Model (DEC)
data("resDEC")  

# Paths to tree and geography files
tree_path <- system.file("extdata", "akodon.new", package = "Herodotools")
phyllip_path <- system.file("extdata", "geo_area_akodon.data", package = "Herodotools")

```


To calculate the Stochastic Mapping you need to have the Biogeographical Model, from which the model parameters will be extracted. You need to inform also, the path to the phylogenetic tree and the phyllip file used in the biogeographical model. Then, the number of stochastic mappings you want to simulate. In this example we are using a small set, only 10 maps. The `calc_bsm()`is wrapping function of the`runBSM()` from `{BioGeoBEARS}`. The output of the function is a list containing the simulated mappings. There are two elements, one with the cladogentic and the other with the anagenetic events. 


```{r calc_bsm }
# calc_bsm ---------------------------------------------------------------------

bsm_result <- calc_bsm(
   BioGeoBEARS.data = resDEC,
   phyllip.file = phyllip_path,
   tree.path = tree_path,
   max.maps = 50,
   n.maps.goal = 10,
   seed = 1234
 )
 
```


## Prepare phylogeny for `{Herodotools}`'s functions

The functions in `{Herodotools}` were designed to map the node in the phylogeny and track changes range form the tip to the root of the phylogeny. To be able to track the changes using the BSM we 
add new nodes to the phylogeny. The new nodes are non-bifurcating nodes that represents a change in the range area state. If the change is due to anagenesis, we use the event time to place it in the daughter branch. If it has a cladogenesis source, we add it on the daughter branch, but very close to the parent node (1e-20).

We have two steps for these additions, first we need to create a data frame for the node insertions and the node area (before insertions). We do it using the functions `get_insert_df()` and `get_bsm_node_area()`. 
    
    

```{r prepare insertions}

# prepare  insertions -------------------------------------------------------
## get_insert_df ----

insert_list <- get_insert_df(
   bsm_result,
   phyllip.file = phyllip_path,
   max.range.size = resDEC$inputs$max_range_size
   )

## get_bsm_node_area ----

node_area_list <- get_bsm_node_area(
  bsm = bsm_result, 
  BioGeoBEARS.data = resDEC,
  phyllip.file = phyllip_path,
  tree.path = tree_path,
  max.range.size = resDEC$inputs$max_range_size)
```


Then we used it in the `insert_nodes()` to insert the nodes to the phylogeny. The `insert_nodes()` functions return a list two elements for each stochastic map:  
    - `$phylo`: a phylogenetic tree with added non-bifurcating nodes  
    - `$node_area`: a data frame with the node area for each added node. If the argument `node_area` is not NULL, all the node area are included in the data frame.   


```{r insert_nodes, message=FALSE, warning=FALSE}
# insert_nodes -----------------------------------------------------------------

bsm_tree <- insert_nodes(
  tree = akodon_newick, 
  inserts = insert_list, 
  node_area = node_area_list)

```

Note that the phylogenies for each BSM output has different number of internal nodes. Also note that there are more internal nodes than tips, which means there are non-bifurcating nodes (or singleton nodes). 

```{r phylo}
bsm_tree[[1]]$phylo
bsm_tree[[10]]$phylo
```

## Run {Herodotools} analyses

Now, we are going to calculate the metrics in {Herodotools}. Before it, let's prepare the data. 
For that we are going to separate the coordinates from the presence-absence data in the `akodon_sites`, filter species present in the phylogeny, and load the evoregions of the sites. 

```{r prep analyses}

# separate coords from presence-absence

site_xy <- akodon_sites %>% 
  dplyr::select(LONG, LAT) 

akodon_pa <- akodon_sites %>% 
  dplyr::select(-LONG, -LAT)

# filter data in communities based on the species in the phylogeny
spp_in_tree <- names(akodon_pa) %in% akodon_newick$tip.label
akodon_pa_tree <- akodon_pa[, spp_in_tree]


# load evoregion and prepare data

evoreg_path <- system.file("extdata", "regions_results.RData", package = "Herodotools")

# load 'regions' object
load(file = evoreg_path)
site_region <- regions$Cluster_Evoregions
evoregion_df <- data.frame(site_xy, site_region)
biogeo_area <- data.frame(biogeo = chartr("12345", "ABCDE", evoregion_df$site_region)) 


# for visualization 
coastline <- rnaturalearth::ne_coastline(returnclass = "sf")
map_limits <- list(
  x = c(-95, -30),
  y = c(-55, 12)
)
```


### Age of arrival

To calculate the metric for each BSM realization we need to build a loop to get the phylogeny and node_area for each BSM and then run the {Herodotools} function, in this case `calc_age_arrival()`. In this example we have only 10 BSM maps, so using `lapply` for the loop is ok in computation time. If you need, you can use some sort of parallel computation (e.g. {future} or {furrr}) to speed the computation time across several BSMs. 


```{r calc_age_arrival}

# calculating age arrival -----
l_age_comm <- lapply(bsm_tree, function(bsm_map){
  
  tree <- bsm_map$phylo
  #tree$node.label <- NULL
  anc_area <- bsm_map$node_area %>% as.matrix()
  
  Herodotools::calc_age_arrival(W = akodon_pa_tree, 
                                tree = tree, 
                                ancestral.area = anc_area, 
                                biogeo = biogeo_area) 
  
})

# Organize results
age_bsm_mtx <- sapply(
  l_age_comm, 
  function(x) x$mean_age_per_assemblage$mean_age_arrival
  )

# summarize results

bsm_metrics <- cbind(
  evoregion_df, 
  age_bsm_mean = rowMeans(age_bsm_mtx),
  age_bsm_sd = apply(age_bsm_mtx, 1, sd)
)

```


### In situ diversidication 

Same logic applies to in situ diversification metrics from `calc_insitu_diversification()`. We create a loop to calculate the in situ diversification for each BSM map.  


```{r calc_insitu_diversification}
# calculating in situ diversification -----

l_div_insitu <- lapply(bsm_tree, function(bsm_map){
  
  tree <- bsm_map$phylo
  #tree$node.label <- NULL
  anc_area <- bsm_map$node_area 
  
  calc_insitu_diversification(
    W = as.matrix(akodon_pa_tree),
    tree = tree, 
    ancestral.area = anc_area, 
    biogeo = biogeo_area, 
    type = "equal.splits"
  )
  
})
```

`calc_insitu_diversification()` returns three community mean metrics, DRjetz, DRinsitu and DRprop. DRjetz is the original DR metric, which does not rely on biogeographical recontructions. DRinsitu is the in situ diversification rate based on biogeographical recostruction. And DRprop is a relationship between the two metrics, quantifying the proportion of diversification that occurred in situ. 

The code below extract each of the community metrics and summarize it for visualization. 


```{r extract calc_insitu_diversification}


# jetz_comm_mean ----
DR_jetz_mtx <- sapply(l_div_insitu, function(x) x$jetz_comm_mean)

# DR Jetz in not depended on the BSM models, so it should not change
max(apply(DR_jetz_mtx, 1, sd))
# SD across BSM models shows max sd as 0. It does not change across BSMs.

# insitu_comm_mean ----
DR_insitu_mtx <- sapply(l_div_insitu, function(x) x$insitu_comm_mean)

# prop_comm_mean ----
DR_prop_mtx <- sapply(l_div_insitu, function(x) x$prop_comm_mean)

# summarize results

bsm_metrics <- cbind(
  bsm_metrics, 
  DR_jetz = DR_jetz_mtx[,1],
  DR_insitu_bsm_mean = rowMeans(DR_insitu_mtx),
  DR_insitu_bsm_sd = apply(DR_insitu_mtx, 1, sd),
  DR_prop_bsm_mean = rowMeans(DR_prop_mtx),
  DR_prop_bsm_sd = apply(DR_prop_mtx, 1, sd)
  )


```


## Visualization of BSM results 

Now we can plot the maps as see the spatial patterns. We are creating a common theme for the maps first, then creating a map for each metric. Finally, we create figures with mean estimates across BSM metrics and uncertainty around those mean estimates. 

```{r viz theme}
# Visualization ----

# create a theme
theme_htools <- list(
  ggplot2::geom_sf(data = coastline, size = 0.4),
  ggplot2::coord_sf(xlim = map_limits$x, ylim = map_limits$y),
  scale_x_continuous(n.breaks = 2),
  scale_y_continuous(n.breaks = 2),
  ggplot2::ggtitle(""),
  ggplot2::theme_bw(),
  ggplot2::guides(fill = ggplot2::guide_colorbar(barheight = unit(2.3, units = "mm"),  
                                        direction = "horizontal",
                                        ticks.colour = "grey20",
                                        title.position = "top",
                                        label.position = "bottom",
                                        title.hjust = 0.5)),
  ggplot2::theme(
    legend.position = "bottom",
    plot.margin = unit(c(0.1, 0.1, 0.1, 0.1), "mm"),
    text = element_text(size = 8)
  )
)
```

### Mean across BSMs

```{r viz mean}

# Spatial patterns, mean across BSMs ----

# Colonization age
plot_age_mean <- bsm_metrics %>% 
  ggplot2::ggplot() + 
  ggplot2::geom_raster(ggplot2::aes(x = LONG, y = LAT, fill = age_bsm_mean)) + 
  rcartocolor::scale_fill_carto_c(type = "quantitative", 
                                  palette = "SunsetDark",
                                  direction = 1) +
  ggplot2::labs(fill = "Mean colonization age (Myr)") +
  theme_htools

# DR jetz
plot_DR_jetz <- bsm_metrics %>% 
  ggplot2::ggplot() + 
  ggplot2::geom_raster(ggplot2::aes(x = LONG, y = LAT, fill = DR_jetz)) + 
  rcartocolor::scale_fill_carto_c(type = "quantitative", 
                                  palette = "SunsetDark",
                                  direction = 1) +
  ggplot2::labs(fill = "DR jetz") +
  theme_htools

# DR in situ
plot_DR_insitu_mean <- bsm_metrics %>% 
  ggplot2::ggplot() + 
  ggplot2::geom_raster(ggplot2::aes(x = LONG, y = LAT, fill = DR_insitu_bsm_mean)) + 
  rcartocolor::scale_fill_carto_c(type = "quantitative", 
                                  palette = "SunsetDark",
                                  direction = 1) +
  ggplot2::labs(fill = "Mean DR in situ") +
  theme_htools

# DR prop
plot_DR_prop_mean <- bsm_metrics %>% 
  ggplot2::ggplot() + 
  ggplot2::geom_raster(ggplot2::aes(x = LONG, y = LAT, fill = DR_prop_bsm_mean)) + 
  rcartocolor::scale_fill_carto_c(type = "quantitative", 
                                  palette = "SunsetDark",
                                  direction = 1) +
  ggplot2::labs(fill = "Mean DR prop") +
  theme_htools
```

```{r, fig.height= 3.5, fig.width=7}

l_plot_bsm_mean <- list(plot_age_mean, plot_DR_jetz, plot_DR_insitu_mean, plot_DR_prop_mean)

patchwork::wrap_plots(l_plot_bsm_mean, nrow = 1, axes = "collect") +
  patchwork::plot_annotation(title = "Mean across BSMs (Except DR jetz)")



```

### Uncertainty across BSMs
```{r viz sd}

# Spatial patterns, uncertainty across BSMs ----


# Colonization age
plot_age_sd <- bsm_metrics %>% 
  ggplot2::ggplot() + 
  ggplot2::geom_raster(ggplot2::aes(x = LONG, y = LAT, fill = age_bsm_sd)) + 
  rcartocolor::scale_fill_carto_c(type = "quantitative", 
                                  palette = "SunsetDark",
                                  direction = 1) +
  ggplot2::labs(fill = "Sd colonization age (Myr)") +
  theme_htools

# DR in situ
plot_DR_insitu_sd <- bsm_metrics %>% 
  ggplot2::ggplot() + 
  ggplot2::geom_raster(ggplot2::aes(x = LONG, y = LAT, fill = DR_insitu_bsm_sd)) + 
  rcartocolor::scale_fill_carto_c(type = "quantitative", 
                                  palette = "SunsetDark",
                                  direction = 1) +
  ggplot2::labs(fill = "Sd DR in situ") +
  theme_htools

# DR prop
plot_DR_prop_sd <- bsm_metrics %>% 
  ggplot2::ggplot() + 
  ggplot2::geom_raster(ggplot2::aes(x = LONG, y = LAT, fill = DR_prop_bsm_sd)) + 
  rcartocolor::scale_fill_carto_c(type = "quantitative", 
                                  palette = "SunsetDark",
                                  direction = 1) +
  ggplot2::labs(fill = "Sd DR prop") +
  theme_htools


```


```{r, fig.height= 3.5, fig.width=7}

l_plot_bsm_sd <- list(plot_age_sd, plot_DR_insitu_sd, plot_DR_prop_sd)

patchwork::wrap_plots(l_plot_bsm_sd, nrow = 1, axes = "collect") +
  patchwork::plot_annotation(title = "SD across BSMs")



```


Using the BSM approach in the {Herodotools} it is possible to propagate the uncertainty in the estimates of ancestral area reconstructions to the community-level in situ metrics. This is important for reliable estimates and interpretation of the in situ diverisification spatial patterns. 

# References:

1. Dupin, J., Matzke, N. J., Särkinen, T., Knapp, S., Olmstead, R. G., Bohs, L., & Smith, S. D. (2017). Bayesian estimation of the global biogeographical history of the Solanaceae. Journal of Biogeography, 44(4), 887–899. https://doi.org/10.1111/jbi.12898







