---
title: "Introduction to SBEARS - Site Based"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Intro_sbears}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Context

Here we will demonstrate how to fit a model of ancestral area reconstruction called 
    SBEARS (Site Based Estimation of Ancestral Range of Species) 
    [(Duarte et al 2025)](https://academic.oup.com/evolut/article-abstract/79/11/2341/8244149?redirectedFrom=fulltext&login=false). SBEARS allows you to reconstruct ancestral
    ranges at finer spatial resolution (grids) without requiring any *a priori* definition
    of biogeographic regions. The output provides the spatial distribution of 
    ancestral nodes in a user-friendly matrix format that can be used
    for other methods in community ecology and macroecology.
    
In SBEARS, the reconstruction is carried out based on the current distribution of 
    species across the geographic space of species belonging to a monophyletic 
    lineage. This information can be organized as an occurrence/
    presence-absence matrix of sites occupied by species. The reconstruction is
    performed using a phylogenetic tree through a stochastic evolutionary model based
    on a Brownian motion model, implemented in the function `fastAnc` of 
    [phytools](https://cran.r-project.org/web/packages/phytools/index.html) R 
    package. In this model, each site is treated as a binary categorical trait 
    representing the presence (state 1) or absence (state 0) of every species in 
    the site, which is reconstructed at the nodes of a phylogeny. 
    
This process is repeated for all sites (with each site corresponding to a 
    different trait) and the final product is a matrix describing the probability 
    value for each node to occur in each site of the grid. The probability is 
    standardized to account for variation in the range size of nodes. This is the procedure called **Single Site**. There is also 
    another option called **Dispersal Assembly** in which the reconstruction 
    probability also depends on the probability of an ancestor occurring 
    in a site dispersing to a neighboring site, given the geographical distance 
    among sites weighted by a dispersal kernel function.
    
Next, we will show how these two methods can be used in the Herodotools package, how 
    to interpret the results of the `calc_sbears` function, and how to visualize them.
    

## Data 

To perform SBEARS reconstruction, we need a community matrix, the geographical
    coordinates of these communities, and a phylogeny for the species occurring
    in the communities. As an example, we will use the same data used in 
    [Maestri et al (2019)](https://besjournals.onlinelibrary.wiley.com/action/oidcStart?redirectUri=%2Fdoi%2Fabs%2F10.1111%2F2041-210X.13492) and [Duarte et al (2025)](https://academic.oup.com/evolut/article-abstract/79/11/2341/8244149?redirectedFrom=fulltext&login=false) that comprise data on the occurrence and phylogeny of 
    Sigmodontinae rodents. The community matrix corresponds to grids of approximately 
    XX km². 

```{r}
library(Herodotools)
data("comm_sbears")
data("coords_sbears")
data("phy_sbears")
```

## Single site algorithm

To perform the single site algorithm, we can use the `calc_sbears` function with
    the argument `method = "single_site"`. 

```{r}
out_single_site <- calc_sbears(x = comm_sbears, 
                               phy = phy_sbears, 
                               coords = coords_sbears,
                               method = "single_site",
                               compute.node.by.sites = TRUE, 
                               make.node.label = TRUE)
```
The function output is a list or a single matrix depending on the arguments 
    used in the function. If the user sets the argument `compute.node.by.site` 
    as TRUE the function returns the output as a list with two elements: 
    
    - `reconstruction`: a matrix with assemblages in rows and ancestral nodes 
        in columns. The values represent the occupation probability of each
        ancestral node in each assemblage.
        
    - `site_node_composition`: a matrix with assemblage in rows and 
       node names in columns. The values represent the presence of a given
       node in a given assemblage based on the occurrence of current species
       in the assemblages.


## Dispersal Assembly algorithm

Here we have to set up two additional arguments: `w_slope` corresponds to the
   slope parameter used in kernel function, and `min_disp_prob` that corresponds
   to the minimum probability of a node between two cells in geographic space.
   
```{r}
out_disp_assembly <- calc_sbears(x = comm_sbears,
                                 phy = phy_sbears, 
                                 coords = coords_sbears, 
                                 method = "disp_assembly",
                                 w_slope = 5,
                                 min_disp_prob = 0.8, 
                                 compute.node.by.sites = TRUE, 
                                 make.node.label = TRUE)
```

The structure of the output for `calc_sbears` function with dispersal assembly
    algorithm is the same as the one with the `single_site` algorithm

