---
title: "Introduction to SBEARS - Site Based"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Intro_sbears}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Context

Here we will demonstrate how to fit a model of ancestral area reconstruction called 
    SBEARS (Site Based Estimation of Ancestral Range of Species) 
    [(Duarte et al 2025)](https://academic.oup.com/evolut/article-abstract/79/11/2341/8244149?redirectedFrom=fulltext&login=false). SBEARS allows you to reconstruct ancestral
    ranges at finer spatial resolution (grids) without requiring any *a priori* definition
    of biogeographic regions. The output provides the spatial distribution of 
    ancestral nodes in a user-friendly matrix format that can be used
    for other methods in community ecology and macroecology.
    
In SBEARS, the reconstruction is carried out based on the current distribution of 
    species across the geographic space of species belonging to a monophyletic 
    lineage. This information can be organized as an occurrence/
    presence-absence matrix of sites occupied by species. The reconstruction is
    performed using a phylogenetic tree through a stochastic evolutionary model based
    on a Brownian motion model, implemented in the function `fastAnc` of 
    [phytools](https://cran.r-project.org/web/packages/phytools/index.html) R 
    package. In this model, each site is treated as a binary categorical trait 
    representing the presence (state 1) or absence (state 0) of every species in 
    the site, which is reconstructed at the nodes of a phylogeny. 
    
This process is repeated for all sites (with each site corresponding to a 
    different trait) and the final product is a matrix describing the probability 
    value for each node to occur in each site of the grid. The probability is 
    standardized to account for variation in the range size of nodes. This is the procedure called **Single Site**. There is also 
    another option called **Dispersal Assembly** in which the reconstruction 
    probability also depends on the probability of an ancestor occurring 
    in a site dispersing to a neighboring site, given the geographical distance 
    among sites weighted by a dispersal kernel function.
    
Next, we will show how these two methods can be used in the Herodotools package, how 
    to interpret the results of the `calc_sbears` function, and how to visualize them.
    

## Data 

To perform SBEARS reconstruction, we need a community matrix, the geographical
    coordinates of these communities, and a phylogeny for the species occurring
    in the communities. As an example, we will use the same data used in 
    [Maestri et al (2019)](https://besjournals.onlinelibrary.wiley.com/action/oidcStart?redirectUri=%2Fdoi%2Fabs%2F10.1111%2F2041-210X.13492) and [Duarte et al (2025)](https://academic.oup.com/evolut/article-abstract/79/11/2341/8244149?redirectedFrom=fulltext&login=false) that comprise data on the occurrence and phylogeny of 
    Sigmodontinae rodents. The community matrix corresponds to grids of approximately 
    XX km². 

```{r}
library(Herodotools)
data("comm_sbears")
data("coords_sbears")
data("phy_sbears")
```

## Single site algorithm

To perform the single site algorithm, we can use the `calc_sbears` function with
    the argument `method = "single_site"`. 

```{r}
out_single_site <- calc_sbears(x = comm_sbears, 
                               phy = phy_sbears, 
                               coords = coords_sbears,
                               method = "single_site",
                               compute.node.by.sites = TRUE, 
                               make.node.label = TRUE)
```

The `calc_sbears` function also supports parallel computation via the
    \code{\link{future}} package. This is recommended for large community
    matrices, as computation time can otherwise be substantial. To run the
    function in parallel, the user can proceed as follows:
    
```{r}
# setting function to work in parallel according to user settings
library(future)
library(progressr)

# Detect safe max cores
ncores <- future::availableCores()  # dynamic
safety_margin <- 1
workers <- max(1, ncores - safety_margin)

plan(multisession, workers = workers)  
handlers("txtprogressbar")   # terminal progress bar + timing info

out_single_site <- calc_sbears(x = comm_sbears, 
                               phy = phy_sbears, 
                               coords = coords_sbears,
                               method = "single_site",
                               compute.node.by.sites = TRUE, 
                               make.node.label = TRUE)

# going back to sequential plan
future::plan(sequential)

```

The function output is a list or a single matrix depending on the arguments 
    used in the function. If the user sets the argument `compute.node.by.site` 
    as TRUE the function returns the output as a list with two elements: 
    
    - `reconstruction`: a matrix with ancestral nodes in rows and sites 
        in columns. The values represent the occupation probability of each
        ancestral node in each site
        
    - `site_node_composition`: a matrix with assemblage in rows and 
       node names in columns. The values represent the presence of a given
       node in a given assemblage based on the occurrence of current species
       representing that lineage in the sites.


## Dispersal Assembly algorithm

When using sbears with the Dispersal Assembly algorithm, two additional
    arguments must be specified: w_slope, which corresponds to the slope
    parameter used in the kernel function, and min_disp_prob, which defines the
    minimum dispersal probability of a node between two cells in geographic
    space.
   
```{r}
out_disp_assembly <- calc_sbears(x = comm_sbears,
                                 phy = phy_sbears, 
                                 coords = coords_sbears, 
                                 method = "disp_assembly",
                                 w_slope = 5,
                                 min_disp_prob = 0.8, 
                                 compute.node.by.sites = TRUE, 
                                 make.node.label = TRUE)
```

The structure of the output for `calc_sbears` function with dispersal assembly
    algorithm is the same as the one with the `single_site` algorithm
    
## Visualizing ancestral areas 

We can use the objects from `calc_sbears` function to plot the ancestral range of ancestral nodes. 
    
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
# dense to long format
long_out_reconstruction <- 
  as.data.frame(as.table(out_single_site$reconstruction))

# naming data frame
colnames(long_out_reconstruction) <- c("nodes", "site", "prob.occurr")

# joining with coordinates
df_coords_sbears <- 
  coords_sbears |> 
  tibble::rownames_to_column("site")

# joining reconstruction with site coordinates
df_reconstruction_coords <- 
  long_out_reconstruction |> 
  left_join(df_coords_sbears, by = "site")


# for visualization 
coastline <- rnaturalearth::ne_coastline(returnclass = "sf")
map_limits <- list(
  x = c(-95, -30),
  y = c(-55, 12)
)

theme_htools <- list(
  ggplot2::geom_sf(data = coastline, size = 0.4),
  ggplot2::coord_sf(xlim = map_limits$x, ylim = map_limits$y),
  ggplot2::scale_x_continuous(n.breaks = 2),
  ggplot2::scale_y_continuous(n.breaks = 2),
  ggplot2::ggtitle(""),
  ggplot2::theme_bw(),
  ggplot2::guides(fill = ggplot2::guide_colorbar(barheight = ggplot2::unit(2.3, units = "mm"),  
                                        direction = "horizontal",
                                        ticks.colour = "grey20",
                                        title.position = "top",
                                        label.position = "bottom",
                                        title.hjust = 0.5)),
  ggplot2::theme(
    legend.position = "bottom",
    plot.margin = unit(c(0.1, 0.1, 0.1, 0.1), "mm"),
    text = element_text(size = 8)
  )
)

# plotting node 114
plot_Node114 <- 
  df_reconstruction_coords |> 
  filter(nodes == "Node114") |> 
  ggplot2::ggplot() + 
  ggplot2::geom_raster(ggplot2::aes(x = longitude, y = latitude, fill = prob.occurr)) + 
  rcartocolor::scale_fill_carto_c(type = "quantitative", 
                                  palette = "SunsetDark",
                                  direction = 1) +
  ggplot2::labs(fill = "Occurrence probability 
                Node 114", title = "Node 114") +
  theme_htools

# plotting node 180
plot_Node180 <- 
  df_reconstruction_coords |> 
  filter(nodes == "Node180") |> 
  ggplot2::ggplot() + 
  ggplot2::geom_raster(ggplot2::aes(x = longitude, y = latitude, fill = prob.occurr)) + 
  rcartocolor::scale_fill_carto_c(type = "quantitative", 
                                  palette = "SunsetDark",
                                  direction = 1) +
  ggplot2::labs(fill = "Occurrence probability 
                Node 180", title = "Node 180") +
  theme_htools

# plotting node 88
plot_Node88 <- 
  df_reconstruction_coords |> 
  filter(nodes == "Node88") |> 
  ggplot2::ggplot() + 
  ggplot2::geom_raster(ggplot2::aes(x = longitude, y = latitude, fill = prob.occurr)) + 
  rcartocolor::scale_fill_carto_c(type = "quantitative", 
                                  palette = "SunsetDark",
                                  direction = 1) +
  ggplot2::labs(fill = "Occurrence probability 
                Node 88", title = "Node 88") +
  theme_htools

# All plots

l_plot_prob_occ <- list(plot_Node114, plot_Node180, plot_Node88)

patchwork::wrap_plots(l_plot_prob_occ, nrow = 1, axes = "collect") +
  patchwork::plot_annotation(title = "Occurrence probability of Nodes in sites")

```



